# Rmd file for class Reproducible Research Assignment 1



#### Set some defaults and load needed libraries

```{r setoptions,echo=TRUE}
library(lattice)
library(Defaults)
setDefaults('as.Date.character', format = '%Y-%m-%d')

```

#### Set the working directory and Read the input CSV file in the Data/ directory

(File has been uncompressed and is called activity.csv)

```{r setwd_read_csv}

setwd("C:\\Data\\On Line Classes\\Data Science Specializations\\Reproducable Results\\Assesment 1\\")
activity <- read.csv("Data/activity.csv", header = TRUE,
                     na.strings = "NA",
            colClasses= c("integer","Date","integer"))

```

#### In the first part of the analysis, any fields missing values (ie, value of "NA") the row is not used (Missing data is "ignored")


#### Aggregate Steps by day


```{r aggregate_by_date}
steps_by_date <- aggregate(steps ~ date, activity,
                           na.action=na.omit,
                           sum)
```

#### Print Histogram of Steps per day
```{r histogram}
hist(steps_by_date$steps,xlab="Number of steps",breaks=length(unique(steps_by_date$steps)),
     main="Histogram of steps taken per day")
```

#### Calculate the Mean and Median of the steps by day

```{r steps_mean_median}
mean_num_steps <- mean(steps_by_date$steps)
median_num_steps <- median(steps_by_date$steps)
```

The mean number of steps by day is `r sprintf("%.0f",mean_num_steps)` and the median number of steps by day is `r sprintf("%.0f",median_num_steps)`.

#### Find the average(mean) of the number of steps in the intervals

```{r aggreagte_by_time_interval}
steps_by_interval <- aggregate(steps ~ interval, activity,
                           na.action=na.omit,
                           mean)

```

#### Plot the average(mean) number of steps per interval

```{r plot_stpes_per_interval}
plot(steps_by_interval$interval, steps_by_interval$steps, type ='l',xlab="Number of steps",ylab="Interval",main="Steps per interval")

```

#### Determine which interval has the most steps on average

```{r max-steps_per_interval}
interval_max_steps <- steps_by_interval[max(steps_by_interval$steps),]$interval

```

The interval with the highest average number of steps was `r interval_max_steps`.


#### Determine how many rows have missing (i.e. "NA") values

```{r count_missing_data}
bad_rows <- sum(is.na(activity))
```

The number of rows with missing data is `r bad_rows`.

#### To make the data "cleaner" we will use the average(mean) number of steps per interval (already calculated above) to fill in any missing data

```{r copy_and_clean}
new_activity <- activity
new_activity$steps <- ifelse((!is.na(new_activity$steps)),new_activity$steps,steps_by_interval[steps_by_interval$interval ==new_activity$interval,]$steps)
```


```{r aggregate_by_date_filled_in_missing_data}
steps_by_date_filled <- aggregate(steps ~ date, new_activity,
                           sum)
```

#### Print Histogram of Steps per day with missing data filled in
```{r histogram_filled_in_missing_data}
hist(steps_by_date_filled$steps,xlab="Number of steps",breaks=length(unique(steps_by_date_filled$steps)),
     main="Histogram of steps taken per day (Missing Data filled in)")
```

#### Calculate the Mean and Median of the steps by day with filled in data

```{r steps_mean_median_filled}
mean_num_steps_filled <- mean(steps_by_date_filled$steps)
median_num_steps_filled <- median(steps_by_date_filled$steps)
```

The mean number of steps by day is `r sprintf("%.0f",mean_num_steps_filled)` and the median number of steps by day is `r sprintf("%.0f",median_num_steps_filled)`.


#### Check for a difference in the average(mean) and the median between the data with and without NA being filled in

```{r check_the difference}
line1 <- c("There ","is not","a difference in the ","median")
line2 <- c("The difference is ","0000"," steps.")
if (mean_num_steps_filled == mean_num_steps) {
        line1[2] <- "is not"
        line1[4] <- "mean."
        mean_line <- paste(line1[1],line1[2],line1[3],line1[4])
} else {
        line1[2] <- "is"
        line1[4] <- "mean"
        line2[2] <- sprintf("%.4f",mean_num_steps_filled - mean_num_steps)
        mean_line <- paste(line1[1],line1[2],line1[3],line1[4],line2[1],line2[2],line2[3])
}

if (median_num_steps_filled == median_num_steps) {
        line1[2] <- "is not"
        line1[4] <- "median."
        median_line <- paste(line1[1],line1[2],line1[3],line1[4])
} else {
        line1[2] <- "is"
        line1[4] <- "median"
        line2[2] <- sprintf("%.4f",median_num_steps_filled - median_num_steps)
        median_line <- paste(line1[1],line1[2],line1[3],line1[4],line2[1],line2[2],line2[3])
}        
```


`r mean_line`

`r median_line`

#### Determine if the day is a Weekday or Weekend and then plot the difference in step per interval by Weekday / Weekend

```{r determine_day_type}
new_activity$weekend <- as.factor(ifelse(weekdays(new_activity$date) %in% c("Saturday","Sunday"),"weekend","weekday"))
xyplot(steps~interval | weekend, data= new_activity,type="l",ylab="Number of steps",layout =c(1,2))

```

